Modal Abstractions for Virtualizing Memory Addresses

### 2024-07-02 examples.tex
+

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ entry \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} }_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ entry \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}   }_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ entry \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}  }_{\rtv}$

@@

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ entry \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val'}  }_{\rtv}$
$\specline{ \sumapaces\rtv\delta  \ast \mathsf{entry\_val.pfn} \in \Xi \setminus \{ entry \} }_{\rtv}$

@@

Concretely speaking, before Line 7 in Figure \ref{fig:calltopteinitialize}, to read the value referenced by physical address \textsf{entry}, first, we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and then obtain the virtual mapping out of the identity mapping $\Xi$ with the side condition (*)
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
assuring that looking up identity mappings are safe. The subtle justification relies on the fact that the full ownership asserts the non-presence of the entry, accessing to which requires allocation a new physical page and having the presence bit set. Acquiring the identity mapping for \textsf{entry} needed before Line 7 in Figure \ref{fig:calltopteinitialize}, we need the knowledge $\textsf{entry} \in \Xi$ to peel-out the fractional token representing the ghost address translation from the offsetted virtual address (\textsf{entry+KERNBASE}) to the physical page address (\textsf{entry}) and physical pointsto relation from the physical address \textsf{entry} to an existentially quantified value (\textsf{entry\_val}) and the side condition (*) we explained above. In in Figure \ref{fig:lookingupidpampping}, we see that all the logical resources needed to construct a virtual-pte pointsto
\[ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \]
to execute Line 7 in \ref{fig:calltopteinitialize} that can either be mutated or not based on the side condition (*).
 \begin{figure*}
   \footnotesize
   \[
 \begin{array}{l}
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast \textsf{entry} \in \Xi }_{\rtv}  \\
\sqsubseteq \\
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{entry\},m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast  \textsf{entry} \notin \Xi }_{\rtv} \\
 \specline{     \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv} \\
\mathsf{mov} \;   [ \mathsf{r8} ],\mathsf{rdi}\\
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{\textsf{entry}\},m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast  \textsf{entry} \notin \Xi}_{\rtv} \\
\specline{  \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast  \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
 \end{array}
 \]
 \vspace{-1em}
\caption{Peel-out the resources needed for the virtual-pte pointsto relation for physical address \textsf{entry} to be accessed in Line 7 in Figure \ref{fig:calltopteinitialize}. }
  \label{fig:lookingupidpampping}
 \end{figure*}

@@

This actually constitutes a very critical piece of the full page table walk verification, and we have verified the critical step for a small x86-64 kernel, which is the physical-to-virtual conversion (often appearing as a macro \texttt{P2V} inside the macro \textsf{ PTE\_PFN\_TO\_ADDR} in C source code). In our small kernel (Line 46 in Figure \ref{calltopteinitialize}), as in larger kernels, \texttt{P2V} is actually just addition, but the correctness of this simple instruction is quite subtle and relies on identity mappings explained in the previous section (Section \ref{subsec:identitymappings}) as a part of understanding our address-space invariant in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.
\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
...
;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
movabs 0x0,rcx	                187: R_X86_64_64	KERN_BASE
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}  }_{\rtv}$
$\specline{\sumpv\paddr\qfrac\true \ast \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land (\textsf{entry\_present entry\_val}) \urcorner }$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr} \ast  \texttt{rcx} \mapsto_{\textsf{r}} \textsf{KERNBASE}  }_{\rtv}$
add    rcx,rax
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}  }_{\rtv}$
$\specline{\sumpv\paddr\qfrac\true \ast \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land (\textsf{entry\_present entry\_val}) \urcorner }$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr} \ast \texttt{rcx} \mapsto_{\textsf{r}} \textsf{KERNBASE} }_{\rtv}$
$\specline{ \texttt{nextpaddr+KERNBASE} \mapsto_{\textsf{vpte,qfracn}} \textsf{nextpaddr w64\_zero} }_{\rtv}$
...
;;next = (pte_t *) next_virt_addr;
;;clean up the stack and return next
\end{lstlisting}
\vspace{-1em}
\caption{Converting a physical address of a PTE to a virtual address (w/o instruction pointer or flag updates).}
\label{fig:p2v}
\end{figure}
As shown in Figure \ref{fig:p2v}, \lstinline|rcx| holds the offset value \textsf{KERNBASE} which gives us the virtual address (\textsf{nextpaddr+KERNBASE}) to look-up in the identity map ($\Xi$) of the kernel invariant. Assuming the path for \textsf{pte\_initialize} is taken, we can obtain the virtual pte-pointsto for the physical address \textsf{nextpaddr} (Line 13 in Figure \ref{fig:p2v}).

>>>>

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} }_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}   }_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}  }_{\rtv}$

@@

$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \mathsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land \lnot(\textsf{entry\_present entry\_val})  \urcorner }_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val'}  }_{\rtv}$
$\specline{ \sumapaces\rtv\delta   \} }_{\rtv}$

@@

Concretely speaking, before Line 7 in Figure \ref{fig:calltopteinitialize}, to read the value referenced by physical address \textsf{entry}, first, we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and then obtain the virtual mapping out of the identity mapping $\Xi$ with the side condition (*)
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
assuring that looking up identity mappings are safe. The subtle justification relies on the fact that the full ownership asserts the non-presence of the entry, accessing to which requires allocation a new physical page and having the presence bit set. Acquiring the identity mapping for \textsf{entry} needed before Line 7 in Figure \ref{fig:calltopteinitialize}, we need the knowledge $\textsf{entry} \in \Xi$ to peel-out the fractional token representing the ghost address translation from the offsetted virtual address (\textsf{entry+KERNBASE}) to the physical page address (\textsf{entry}) and physical pointsto relation from the physical address \textsf{entry} to an existentially quantified value (\textsf{entry\_val}) and the side condition (*) we explained above. In in Figure \ref{fig:lookingupidpampping}, we see that all the logical resources needed to construct a virtual-pte pointsto
\[ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \]
to execute Line 7 in \ref{fig:calltopteinitialize} that can either be mutated or not based on the side condition (*).
 \begin{figure*}
   \footnotesize
   \[
 \begin{array}{l}
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast \textsf{entry} \in \Xi }_{\rtv}  \\
\sqsubseteq \\
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} }_{\rtv}\\
   \specline{ \sumapaces\rtv\delta  }_{\rtv} \\
\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv} \\
\mathsf{mov} \;   [ \mathsf{r8} ],\mathsf{rdi}\\
 \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} }_{\rtv}\\
   \specline{ \sumapaces\rtv\delta  }_{\rtv} \\
\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv} \\
 \end{array}
 \]
 \vspace{-1em}
\caption{Peel-out the resources needed for the virtual-pte pointsto relation for physical address \textsf{entry} to be accessed in Line 7 in Figure \ref{fig:calltopteinitialize}. }
  \label{fig:lookingupidpampping}
 \end{figure*}

@@

This actually constitutes a very critical piece of the full page table walk verification, and we have verified the critical step for a small x86-64 kernel, which is the physical-to-virtual conversion (often appearing as a macro \texttt{P2V} inside the macro \textsf{ PTE\_PFN\_TO\_ADDR} in C source code). In our small kernel (Line 46 in Figure \ref{calltopteinitialize}), as in larger kernels, \texttt{P2V} is actually just addition, but the correctness of this simple instruction is quite subtle and relies on identity mappings explained in the previous section (Section \ref{subsec:identitymappings}) as a part of understanding our address-space invariant in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.
\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
...
;;uintptr_t next_virt_addr = (uintptr_t) P2V(next<<12);
movabs 0x0,rcx	                187: R_X86_64_64	KERN_BASE
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val'}  }_{\rtv}$
$\specline{ \sumapaces\rtv\delta  \ast \mathsf{entry\_val'.pfn}<<12 \in \Xi \setminus \{ \mathsf{entry} \} }_{\rtv}$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val\_present} \ast \ulcorner  \textsf{qfrac} = 1 \land (\textsf{entry\_present entry\_val'}) \urcorner }_{\rtv}$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr}  }_{\rtv}$
add    rcx,rax
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} , \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val'}  }_{\rtv}$
$\specline{ \sumapaces\rtv\delta  \ast \mathsf{entry\_val'.pfn}<<12 \in \Xi \setminus \{ \mathsf{entry} \} }_{\rtv}$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val\_present} \ast \ulcorner  \textsf{qfrac} = 1 \land (\textsf{entry\_present entry\_val'}) \urcorner }_{\rtv}$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr}  }_{\rtv}$
$\specline{ \texttt{nextpaddr+KERNBASE} \mapsto_{\textsf{vpte,qfracn}} \textsf{nextpaddr w64\_zero} }_{\rtv}$
...
;;next = (pte_t *) next_virt_addr;
;;clean up the stack and return next
\end{lstlisting}
\vspace{-1em}
\caption{Converting a physical address of a PTE to a virtual address (w/o instruction pointer or flag updates).}
\label{fig:p2v}
\end{figure}
As shown in Figure \ref{fig:p2v}, \lstinline|rcx| holds the offset value \textsf{KERNBASE} which gives us the virtual address (\textsf{nextpaddr+KERNBASE}) to look-up in the identity map ($\Xi$) of the kernel invariant. Assuming the path for \textsf{pte\_initialize} is taken, we can obtain the virtual pte-pointsto for the physical address \textsf{nextpaddr} (Line 13 in Figure \ref{fig:p2v}).
